\ourtitle

### 2020-07-02 paper/article.tex
Update paper

\begin{claim}[Total Order]
\label{claim:total-order} To allow potentially infinite sequences to be played in realtime, music events must always be generated in the correct order: otherwise the sequence would have to be generated fully before we could be sure of what event was actually next.
\end{claim}

@@

\paragraph*{Syntax}
While our goal is not to delve into every little detail of the language (because most are just similar to all popular programing languages), and we'd rather focus on how the tight integration of the Music data type into the language changes some aspects of it. Still, for the sake of making it easier to understand this paper, we will briefly discuss the syntactic features provided by the language.

A non-exhaustive list of accepted expressions are:
\begin{description}
    \item[Variables] \verb|$var|
    \item[Function Calls] \verb|function_name( expr, named_arg = expr )|
    \item[Function Declaration] A function declaration is an expression (allowing higher order functions). Function names are optional. Giving a name to a function is equivalent to assigning an anonymous function to a variable with that name.
        \begin{description}
            \item[Anonymous] \verb|$name = fun() {}|
            \item[Arguments] \verb|fun name($arg, ref $ref_arg, optional = expr) {}|
            \item[Single Expression] \verb|fun name() => expr|
            \item[Multiple Statements] \verb|fun name() {}|
        \end{description}

    \item[Arithmetic Operators and Parenthesis Grouping] \verb|( expr + expr ) * expr|
    \item[Literals] \verb|1, true, none, "string", @["array"], @{ "object": 1 }|
    \item[Code Blocks] \verb|{ stmt; stmt; stmt }|
    \item[Object and Array accessors] \verb|$var::property| and \verb|$var::[ $index ]|
    \item[Keyboard Declarations] \verb|@keyboard { a: print("Key a"); b: print("Key b"); }|
\end{description}

A non-exhaustive list of accepted statements are:
\begin{description}
    \item[Conditionals] \verb|if (expr) {stmt} else {stmt}|
    \item[Loops] \verb|for ($var in expr) {stmt}| and \verb|while(expr) {stmt}|
    \item[Variable Declarations] \verb|$var = expr|
    \item[Returns] \verb|return expr|
    \item[Expressions] Any expression can be used anywhere a statement is expected
\end{description}

Statements must always end with \textbf{semicolons}, even loops and conditionals with the exception of the \textbf{last semicolon} in a list of statements, which is optional.

@@

        \textbf{type} \verb|Chord, Music -> Music|
    \item[Transpose] \verb'Music + Integer' \textit{and} \verb'Music - Integer'

@@

\footnotetext{\url{https://drive.google.com/open?id=1IIm8PQkLsNFMK9MNSVubJSG6SP6KwhPL}}

\paragraph*{Parallel}
The parallel operator enables playing multiple sequences of musical events simultaneously. However our events are emitted as a single sequence of ordered events, thus requiring merging the multiple sequences into a single one, while maintaining the properties of laziness and order. The operator assumes that each of its operands already maintains those properties on their own, and so is only in charge of making sure the merged sequence does so as well. With this in mind, it relies on a custom \textit{merge sorted} algorithm for iterables (not related to the most common merge sort algorithm by John von Neumann).

@@

We also make use of the \texttt{stretch} function included in our standard library, which receives two arguments of type \textbf{Music}, and returns the first one but with it's duration proportionally stretched (or shrunk) to have the same total duration as the second musical expression.

\begin{lstlisting}[label={lst:fugue},caption={Example of repeating the same note},captionpos=t,abovecaptionskip=-\medskipamount]
fun fugue ( $subj, $resp ) => 
    ( $subj $resp | stretch( r, $subj ) ( $subj + 7 ) );

@@

$resp = E/2 c/2 B/2 A/2   G/2 F/2 _E/2 D/2   C _e d c
        _B A _B c   ^F G _A F;

@@

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/fugue.png}}%
  }%
  \caption{Generated music sheet for fugue example, audio version available \href{https://drive.google.com/open?id=1dIfvnhhKn73Vpp0W6ss6RLsv6PQ_HFTF}{\underline{here}}\protect\footnotemark.}
  \label{fig:fugue}
\end{figure}
\footnotetext{\url{https://drive.google.com/open?id=1dIfvnhhKn73Vpp0W6ss6RLsv6PQ_HFTF}}

@@

\begin{lstlisting}[caption={Declaring a grid},label=list:8,captionpos=t,abovecaptionskip=-\medskipamount]
$grid = Grid( 1,
    forgiveness_left = 125, # 1/8
    range_left = 375, # 3/8
    forgiveness_right = 125, #1/8
    range_right = 375 # 3/8
);
\end{lstlisting}

>>>>

\begin{claim}[Total Order]
\label{claim:total-order} To allow potentially infinite sequences to be played in realtime, music events must always be generated in the correct order: otherwise the sequence would have to be generated fully before we could be sure of what event was actually next in line to play.
\end{claim}

@@

\paragraph*{Syntax}
% While we will focus more on the musical-related aspects of our language, here is a non-exhaustive list of accepted statements and expressions:
While our goal is not to delve into every little detail of the language (because most are just similar to all popular programing languages), and we'd rather focus on how the tight integration of the Music data type into the language changes some aspects of it. Still, for the sake of making it easier to understand this paper, we will briefly discuss the syntactic features provided by the language.

Here is a non-exhaustive list of accepted expressions:

\begin{table}[h]
\centering
\def\arraystretch{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|}
\hline
\textbf{EXPRESSIONS}       & \textbf{SYNTAX}                                                                         \\ \hline
Variables                  & \verb|$var|                                                                             \\ \hline
Function Calls             & \verb|function_name( expr, named_arg = expr )|                                         \\ \hline
Function Declaration       & \begin{tabular}[r]{@{}r@{}}\verb|$name = fun() {}|\\ \verb|fun name($arg, ref $ref_arg, optional = expr) {}|\\\verb|fun name() => expr|\\\verb|fun name() {}|\end{tabular} \\ \hline
Modules                    & \verb|import "file.mkl"| \\ \hline
Arithmetic Operators       & \verb|a + b|; \verb|a * b|; \verb|a - b|; \verb|a / b|                                                                  \\ \hline
Literals                   & \verb|1, true, none, "string"|                                  \\ \hline
\end{tabular*}
\caption{List of valid expressions in Musikla}
\label{tab:syntax-expressions}
\end{table}

A \textbf{function declaration} is an expression (allowing higher order functions). Function names are optional. Giving a name to a function is equivalent to assigning an anonymous function to a variable with that same name.

\newpage

\begin{table}[h]
\centering
\def\arraystretch{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|}
\hline
\textbf{STATEMENTS}        &  \textbf{SYNTAX} \\ \hline
Attribution                & \verb|$var = expr| \\ \hline
Code Blocks                & \verb|{ stmt; stmt; stmt }| \\ \hline
Conditionals               & \verb|if (expr) {stmt} else {stmt}| \\ \hline
Loops                      & \begin{tabular}[r]{@{}r@{}}\verb|for ($var in expr) {stmt}| \\ \verb|while (expr) {stmt}|\end{tabular} \\ \hline
Returns                    & \verb|return expr| \\ \hline
\end{tabular*}
\caption{List of valid statements in Musikla}
\label{tab:syntax-statements}
\end{table}

Statements must always end with \textbf{semicolons}, even loops and conditionals, with the exception of the \textbf{last semicolon} in a list of statements, which is optional.

\begin{table}[h]
\centering
\def\arraystretch{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|}
\hline
\textbf{LISTS}             & \textbf{SYNTAX} \\ \hline
Declaration                & \verb|@[ "list" ]| \\ \hline
Indexed                    & \verb|$var::[ $index ]| \\ \hline\hline
\textbf{OBJECTS}           &  \\ \hline
Declaration                & \verb|@{ "object": 1 }| \\ \hline
Attributes                 & \verb|$obj::property| \\ \hline
Methods                    & \verb|$obj::method()| \\ \hline
\end{tabular*}
\caption{List and Object operations in Musikla}
\label{tab:syntax-lists-objects}
\end{table}
To avoid ambiguities between chords, lists (square brackets), code blocks, objects (brackets), both list and object literals are prefixed with the \texttt{@} symbol.

Also, we use two colons as separators to access lists and objects' properties and methods, because the period is used in the musical notation for indicating dotted notes, and as such could be confusing and ambiguous sometimes.

\begin{table}[h]
\centering
\def\arraystretch{1.1}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|}
\hline
\textbf{KEYBOARDS}         & \textbf{SYNTAX} \\ \hline
Keyboard Declarations      & \verb|@keyboard {  }| \\ \hline
Keyboard with Modifiers    & \verb|@keyboard hold extend {  }| \\ \hline\hline
\textbf{KEYBOARD BODY}     &  \\ \hline
Expression action          & \verb|key: expr| \\ \hline
Action Block               & \verb|key: { stmt; stmt }| \\ \hline\hline
\textbf{KEYS}\footnotemark &  \\ \hline
Single Keys                & \verb|a, b, 1, 2, ".", up, f12, ctrl a, ctrl shift f| \\ \hline
Composite Keys             & \verb|ctrl a, ctrl shift left| \\ \hline
MIDI Keys                  & \verb|[c] [D,] [a'']| \\ \hline
\end{tabular*}
\caption{Keyboard declaration and body syntax}
\label{tab:syntax-keyboards}
\end{table}
\footnotetext{Examples of the types of keys accepted. Rules for the key syntax are described below the table.}
A keyboard can have modifiers (such as \textbf{hold} to start/stop the music when the key is pressed/released; \textbf{toggle} to start/stop the music when the key is pressed; \textbf{extend} to override the duration of the notes to last as long as the key is active;).

Keys can be any alphanumeric character, or any character wrapped in quotes. We can also create key combinations with the modifiers \textbf{ctrl}, \textbf{shift} and \textbf{alt}. Notes can be used as keys (to program MIDI keyboards connected to the computer) by wrapping them in square brackets. Each key can have an expression associated, or a code block with a list of arbitrary statements.

\begin{table}[h]
\centering
\def\arraystretch{1.2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}|l|r|}
\hline
\textbf{MUSIC}             & \textbf{SYNTAX} \\ \hline
Notes                      & \verb|C, C c c' A2/3 _A ^A| \\ \hline
Chords                     & \verb|[CFG] [Cm]/2| \\ \hline
Rests                      & \verb|r r1/2| \\ \hline
Modifiers                  & \verb|S4/4 T60 L/2 V70 O2 I2| \\ \hline
Parallel                   & \texttt{C E | F A} \\ \hline
\end{tabular*}
\caption{Music literal expressions}
\label{tab:syntax-music}
\end{table}
A note's octave is expressed by the case of the note (uppercase and commas mean lower pitch, lowercase and single quotes mean higher pitch). They can be prefixed by an optional accidental (underscore for flats, caret for sharps). They can be suffixed by an optional note length (either an integer or a fraction). Chords can list each of their notes, or just the root note and a suffix (\textbf{CM} for \textit{C major} and \textbf{Cm} for \textit{C minor}, for example).

@@

        \textbf{type} \verb|Chord, Music -> Music|
    \item[Retime] \verb'Music ** MusicOrLength'

        \textbf{type} \verb|Music, Union[Music, Float] -> Music|

        Stretches (or shrinks) proportionally the length of the first music expression to be the same as the length of the second.
    \item[Transpose] \verb'Music + Integer' \textit{and} \verb'Music - Integer'

@@

\footnotetext{\url{https://drive.google.com/open?id=1IIm8PQkLsNFMK9MNSVubJSG6SP6KwhPL}}
\newpage
\paragraph*{Parallel}
The parallel operator enables playing multiple sequences of musical events simultaneously. However our events are emitted as a single sequence of ordered events, thus requiring merging the multiple sequences into a single one, while maintaining the properties of laziness and order. The operator assumes that each of its operands already maintains those properties on their own, and so is only in charge of making sure the merged sequence does so as well. With this in mind, it relies on a custom \textit{merge sorted} algorithm for iterables (not related to the most common merge sort algorithm by John von Neumann).

@@

\begin{lstlisting}[label={lst:fugue},caption={Example of repeating the same note},captionpos=t,abovecaptionskip=-\medskipamount]
fun fugue ( $subj, $resp ) => 
    ( $subj $resp | r ** $subj ( $subj + 7 ) );

@@

$resp = _E/2 c/2 B/2 A/2   G/2 F/2 _E/2 D/2   C _e d c
        _B A _B c   ^F G A ^F;

@@

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/fugue.png}}%
  }%
  \caption{Generated music sheet for fugue example, audio version available \href{https://drive.google.com/file/d/1dIfvnhhKn73Vpp0W6ss6RLsv6PQ_HFTF/view}{\underline{here}}\protect\footnotemark.}
  \label{fig:fugue}
\end{figure}
\footnotetext{\url{https://drive.google.com/file/d/1dIfvnhhKn73Vpp0W6ss6RLsv6PQ_HFTF/view}}

@@

\begin{lstlisting}[caption={Declaring a grid with some additional optional parameters},label=list:8,captionpos=t,abovecaptionskip=-\medskipamount]
$grid = Grid( 1,
    forgiveness_left = 125,  # 1/8
    range_left = 375,        # 3/8
    forgiveness_right = 125, # 1/8
    range_right = 375        # 3/8
);
\end{lstlisting}
